---
title: "Rawan's CSSS508 notes for R programming"
author: "Rawan Alaseeri"
date: "2025-02-20"
output: github_document
---

A lot of the things in this document are links and things I found useful while learning R programming, mainly from [CSSS508](https://clanfear.github.io/CSSS508/), thanks Chuck!

It is very important through learning R is to annotate your code with commenting and with R markdown..

## R Markdown

```{r Markdown, eval=FALSE}
# dont show code output -> echo=FALSE
# shortcut for creating a chunk -> ctrl+Alt+I
```

Links:

1.  [Introduction to R Markdown](https://rmarkdown.rstudio.com/lesson-1.html)

2.  [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)

3.  [Cheat Sheets](https://posit.co/resources/cheatsheets/)

4.  [R Colors Sheet](https://sites.stat.columbia.edu/tzheng/files/Rcolor.pdf)

5.  [PrettyDoc](https://prettydoc.statr.me/index.html) for R markdown themes

6.  [Posit R Markdown Cheat Sheet](https://rstudio.github.io/cheatsheets/html/rmarkdown.html)

7.  [R Studio IDE Cheat Sheet](https://rstudio.github.io/cheatsheets/html/rstudio-ide.html)

### Couple more housekeeping things

```{r, eval=FALSE}
# save R objects using 
save(new.object, file="new_object.RData")
# then you can open it using
load(new_object.RData)
# find working directory with 
getwd() # mine is in downloads which is probably not the best but eh
# to change the working directory 
# setwd("") #my computer copies the file path as \ so I should change it to /

# install and load packages
install.packages()
library()
```

Note to myself: for the love of God, manage your files and set folders and a consistent folder. Also, a good way to manage files in R studio is to create a project (eg. Thesis, Learning R, etc). this is nice because it also changes your working directory if you need to

`.Rmd` is used to produce **document**

`.R` is just for running things

`.html` are output docs

### Subsetting

```{r, eval=FALSE}
# check the structure of data using
str()
# check all catagories in some data
unique()
```

### `dplyr`

you cane use the `magritter` opration to pip data between functions. magritter allows you to read or write the code in a more intuitive way (from left to right instead from inside out)

```{r, eval=FALSE}
# and then %>%
```

filter data frames. For example:

```{r, eval=FALSE}
data %>% filter(Country == "Algeria")
# this reads as: from data, filter the data where the country is Algeria
# use == instread of = 
```

More about [operators](https://www.datacamp.com/doc/r/operators)

For example:

-   `!=`: not equal to

-   `>`, `>=`, `<`, `<=`: less than, less than or equal to, etc.

-   `%in%`: used with checking equal to one of several values

--

Or we can combine multiple logical conditions:

-   `&`: both conditions need to hold (AND)

-   `|`: at least one condition needs to hold (OR)

-   `!`: inverts a logical condition (`TRUE` becomes `FALSE`, `FALSE` becomes `TRUE`)

## Visualising Data with `ggplot2`

`ggplot2` graphics objects consist of two primary components:

1.  **Layers**, the components of a graph.

    -   We *add* layers to a `ggplot2` object using `+`.
    -   This includes lines, shapes, and text.

2.  **Aesthetics**, which determine how the layers appear.

    -   We *set* aesthetics using *arguments* (e.g. `color="red"`) inside layer functions.
    -   This includes locations, colors, and sizes.
    -   Aesthetics also determine how data *map* to appearances.

**Layers** are the components of the graph, such as:

-   `ggplot()`: initializes `ggplot2` object, specifies input data
-   `geom_point()`: layer of scatterplot points
-   `geom_line()`: layer of lines
-   `ggtitle()`, `xlab()`, `ylab()`: layers of labels
-   `facet_wrap()`: layer creating separate panels stratified by some factor wrapping around
-   `facet_grid()`: same idea, but can split by two variables along rows and columns (e.g. `facet_grid(gender ~ age_group)`)
-   `theme_bw()`: replace default gray background with black-and-white

Layers are separated by a `+` sign. For clarity, I usually put each layer on a new line, unless it takes few or no arguments (e.g. `xlab()`, `ylab()`, `theme_bw()`).

**Aesthetics** control the appearance of the layers:

-   `x`, `y`: $x$ and $y$ coordinate values to use

-   `color`: set color of elements based on some data value

-   `group`: describe which points are conceptually grouped together for the plot (often used with lines)

-   `size`: set size of points/lines based on some data value

-   `alpha`: set transparency based on some data value

-   Arguments inside `aes()` (**mapping aesthetics**) will *depend on the data*, e.g. `geom_point(aes(color = continent))`.

Example of a plot syntax:

```{r, eval=FALSE}
lifeExp_by_year <- 
  ggplot(data = gapminder, 
       aes(x = year, y = lifeExp, 
           group = country, 
           color = continent)) +
  geom_line() +
  xlab("Year") + 
  ylab("Life expectancy") +
  ggtitle("Life expectancy over time") +
  theme_bw() + 
  facet_wrap(~ continent) +
  theme(legend.position = "none")
```

this will save the plot into lifeExp_by_year in case i need to use it again.. this saves the code of the plot which helps you add and change things in the code for the plot

### Scales for Color, Shape, etc.

**Scales** are layers that control how the mapped aesthetics appear. You can modify these with a `scale_[aesthetic]_[option]()` layer where `[aesthetic]` is `color`, `shape`, `linetype`, `alpha`, `size`, `fill`, etc. and `[option]` is something like `manual`, `continuous` or `discrete` (depending on nature of the variable).

Examples:

-   `scale_linetype_manual()`: manually specify the linetype for each different value

-   `scale_alpha_continuous()`: varies transparency over a continuous range

-   `scale_color_brewer(palette = "Spectral")`: uses a palette from <http://colorbrewer2.org> (great site for picking nice plot colors!)

### Saving `ggplot` Plots

When you knit an R Markdown file, any plots you make are automatically saved in the "figure" folder in `.png` format. If you want to save another copy (perhaps of a different file type for use in a manuscript), use `ggsave()`:

```{r, eval=FALSE}
ggsave("I_saved_a_file.pdf", plot = lifeExp_by_year,
       height = 3, width = 5, units = "in")
```

If you didn't manually set font sizes, these will usually come out at a reasonable size given the dimensions of your output file.

## dplyr

### piping

```{r, eval=FALSE}
take_these_data %>%
    do_first_thing(with = this_value) %>%
    do_next_thing(using = that_value) %>% ...
```

### Another Operator: `%in%`

Common use case: Filter rows to things in some *set*.

We can use `%in%` like `==` but for matching *any element* in the vector on its right<sup>1</sup>.

```{r, eval=FALSE}
former_yugoslavia <- c("Bosnia and Herzegovina", "Croatia", 
                       "Montenegro", "Serbia", "Slovenia")
yugoslavia <- gapminder %>% filter(country %in% former_yugoslavia)
tail(yugoslavia, 2)
```

### `distinct()`

You can see all the *unique values* in your data for combinations of columns using `distinct()`:

```{r, eval=FALSE}
gapminder %>% distinct(continent, year)
```

### Sorting: `arrange()`

Along with filtering the data to see certain rows, we might want to sort it:

```{r, eval=FALSE}
yugoslavia %>% arrange(year, desc(pop))
```

The data are sorted by ascending `year` and descending `pop`.

### Keeping Columns: `select()`

Not only can we limit rows, but we can include specific columns (and put them in the order listed) using `select()`.

```{r, eval=FALSE}
yugoslavia %>% select(country, year, pop) %>% head(4)
```

We can instead drop only specific columns with `select()` using `-` signs:

```{r}
yugoslavia %>% select(-continent, -pop, -lifeExp) %>% head(4)
```

### Create New Columns: `mutate()`

In `dplyr`, you can add new columns to a data frame using `mutate()`.

```{r, eval=FALSE}
yugoslavia %>% filter(country == "Serbia") %>%
    select(year, pop, lifeExp) %>%
    mutate(pop_million = pop / 1000000,
           life_exp_past_40 = lifeExp - 40) %>%
    head(5)
```

Note you can create multiple variables in a single `mutate()` call by separating the expressions with commas.

### `case_when()`

`case_when()` performs multiple `ifelse()` operations at the same time. `case_when()` allows you to create a new variable with values based on multiple logical statements. This is useful for making categorical variables or variables from combinations of other variables.

```{r, eval=FALSE}
gapminder %>% 
  mutate(gdpPercap_ordinal = 
    case_when(
      gdpPercap <  700 ~ "low",
      gdpPercap >= 700 & gdpPercap < 800 ~ "moderate",
      TRUE ~ "high" )) %>% # Value when all other statements are FALSE
  slice(6:9) # get rows 6 through 9
```

### `pull()`

Sometimes you want to extract a single column from a data frame as a *vector* (or single value).

`pull()` *pulls* a column of a data frame out as a vector.

```{r, eval=FALSE}
gapminder %>% pull(lifeExp) %>% head(4)
```

```{r, eval=FALSE}
gapminder %>% select(lifeExp) %>% head(4)
```

### Join Types: Rows and columns kept

There are many types of joins<sup>1</sup>...

-   `A %>% left_join(B)`: keep all rows from `A`, matched with `B` wherever possible (`NA` when not), keep columns from both `A` and `B`

-   `A %>% right_join(B)`: keep all rows from `B`, matched with `A` wherever possible (`NA` when not), keep columns from both `A` and `B`

-   `A %>% inner_join(B)`: keep only rows from `A` and `B` that match, keep columns from both `A` and `B`

-   `A %>% full_join(B)`: keep all rows from both `A` and `B`, matched wherever possible (`NA` when not), keep columns from both `A` and `B`

-   `A %>% semi_join(B)`: keep rows from `A` that match rows in `B`, keep columns from only `A`

-   `A %>% anti_join(B)`: keep rows from `A` that *don't* match a row in `B`, keep columns from only `A`

### Matching Criteria

We say rows should *match* because they have some columns containing the same value. We list these in a `by =` argument to the join.

Matching Behavior:

-   No `by`: Match using all variables in `A` and `B` that have identical names
-   `by = c("var1", "var2", "var3")`: Match on identical values of `var1`, `var2`, and `var3` in both `A` and `B`
-   `by = c("Avar1" = "Bvar1", "Avar2" = "Bvar2")`: Match identical values of `Avar1` variable in `A` to `Bvar1` variable in `B`, and `Avar2` variable in `A` to `Bvar2` variable in `B`

Note: If there are multiple matches, you'll get *one row for each possible combination* (except with `semi_join()` and `anti_join()`).

Need to get more complicated? Break it into multiple operations.
